An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging

### 2018-10-11 HowtoHowtoCapon/document.tex
Testing change of frame letters

This article is outlined as follows: In Section \ref{methods}, we introduce the concept of adaptive beamforming and provide details on the MVDR method. Then, in \ref{maptogpu}, we investigate the complexity issue, discuss means for reducing arithmetic complexity, and detail an implementation that makes efficient use of the GPU's parallel resources. The final design is assessed in Sections \ref{images_and_benchmarks} and \ref{discussion}, where we provide benchmarks, comparisons with similar CPU implementations, and measures of how efficiently our implementation makes use of the GPU's resources.

@@

\section{Methods}\label{methods}

@@

\subsection{Beamforming}
\ifPhdDoc
\begin{figure}[htbp]\centering
\includegraphics[width=\linewidth]{gfx/buske1\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[htbp]\centering
\includegraphics[width=0.8\linewidth]{gfx/buske1\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/beamforming.eps}
\fi\fi%
\caption{Beamforming principle. Signal signature is first removed by matched filtering. Then, before summation, a suitable set of delays $\Delta$ and weights $w$ are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{beamforming}
\end{figure}
Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter (Fig. \ref{beamforming}). Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$. To simplify notation, we make the dependence on $\theta$ implicit from now on.

By definition, the beamformer output $z[n]$ can now be expressed as the weighted sum of all the delayed data samples:
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
\end{align}
where $w_m$ is the weight factor assigned to channel $m$. With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer. A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.

Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield. In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality. The MVDR beamformer is one such method. It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}. This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{weights}
\end{gather}
where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. Since we pre-steer our data to every pixel in the image we simplify (\ref{weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering. To estimate $\R$ we compute a sample covariance matrix $\eR$. In this computation we perform some degree of:
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
\end{itemize}%
%
% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009a}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}. 
%
Combined, these steps will also ensure a numerically well conditioned $\eR$.

We perform \emph{temporal} and \emph{spatial averaging} first and put the result in an intermediate sample covariance matrix $\breve{\R}$. To do this we need to segment our array into subarrays. If we let $x_l[n]$ represent the data vector from subarray $l$ by
\begin{gather}
\x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T,
\end{gather}
then $\breve{\R}$ can be calculated as
\begin{gather}
\breve{\R}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n']\x_l\H[n'] \in\mathbb{C}^{L,L},\label{spatialR}
\end{gather}
where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.

The final estimate $\eR$ is found by adding a fraction $d$ of the total power of $\breve{\R}[n]$ to its diagonal~\cite{Synnevag2007}:
\begin{align}
\eR[n] = \breve{\R}[n] + \I \frac{d}{L} \tr\{\breve{\R}[n]\},\label{finalR}
\end{align}
where $\I$ is an identity matrix, $\tr\{\cdot\}$ represents the matrix trace operation, and $\tr\{\breve{\R}[n]\}$ is an estimate of the energy received from this pixel.

Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{M,M}$ to $\mathbb{C}^{L,L}$, and hence will produce an $L$-element weight set when substituted into (\ref{z}). This weight set is applied to all the subarrays, before computing the beamformer output as in (\ref{z}). Or, equivalently, we may apply the weight set to the sum of all the subarrays by
%
\begin{align}
z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{finalZ}
\end{align}
%
\ifPhdDoc
\begin{figure}[tbp]\centering
\includegraphics[width=\linewidth]{gfx/buske2\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske2\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/implementation.eps}
\fi\fi%
\caption{MVDR beamforming. For each pixel in range and azimuth:\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed; \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver;\newline
3. and the beamformer output $z$ is computed from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
\end{figure}
As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
\begin{enumerate}
\item computing the sample covariance matrix $\eR$ in (\ref{finalR});
\item computing $\eRi\1$ in (\ref{weights});
\item computing the beamformer output $z$ in (\ref{finalZ}).
\end{enumerate}
Next we will evaluate these steps in terms of arithmetic complexity, and then discuss their mappability to parallel hardware.

@@

If we neglect spatial and temporal averaging, then the computation of $\eR$ is reduced to a single outer product with complexity of O($M^2$), and the inversion is then of O($M^3$). This might lead us to believe that the inversion step is by far the most complex. But if we implement spatial and temporal averaging as in (\ref{spatialR}), then computing $\eR$ is of O($N_K N_L L^2$) and the inversion of O($L^3$). Computing $\eR$ is now the most complex operation whenever $N_LN_K>L$.
%
\ifPhdDoc
\begin{figure}[tbp]\centering
\includegraphics[width=\linewidth]{gfx/buske3\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske3\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_complexity.eps}
\fi\fi%
\caption{Per-pixel computational complexity of the steps in MVDR beamforming (before any optimizations). To avoid signal cancellation in an active sonar system we usually set $L<\frac{M}{2}$, in which region the computation of $\eR$ dominates in terms of arithmetic complexity, especially when performing temporal averaging.}\label{mvdr_complexity}
\end{figure}
To visualize these relations, and include the effects of using complex numbers, we set up complexity formulas that account for the total number of arithmetic operations for each step in the MVDR process (see appendix \ref{mvdr_formulas}). The only excluded operation was partial pivoting used in the inversion step, but this should contribute marginally to the end result. The entire range of possible subarray sizes from $L\in[1,M]$ was finally evaluated, with temporal averaging set to $K\in\{0,1,2\}$, and the number of channels set to $M\in\{8,16,32\}$. The results are shown in \Fig{mvdr_complexity}.

@@

\section{Mapping the MVDR to a GPU}\label{maptogpu}

@@

To use Nvidia's own terminology, the Quadro 6000 comprises 14 streaming multiprocessors (SMs), each having 32 CUDA cores that execute a common program called a kernel. Combined these cores deliver a peak performance of more than 1\;Tflop/s (appendix \ref{throughput}). In practice this performance is hard to obtain, but one can get fairly close by balancing the load evenly on all cores, and by trying to avoid that some cores are forced to idle due to a pending data transfer or thread synchronization. As the steps in the MVDR method require different strategies for achieving this, we have designed a different kernel and configuration for each of them. We will pay particular attention to building $\eR$, since the potential for gaining overall speedups are greatest here.

@@

\ifPhdDoc
\begin{figure}[tbp]\centering
\includegraphics[width=\linewidth]{gfx/buske4\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske4\figPostfix.eps}
\else
\begin{figure}[t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_build_R.eps}
\fi\fi%
\caption{Step 1: Building $\eR$. This is a visualization of how $\eR$ could be built in a case with $M=5$ sensors, with subarray size $L=3$ and temporal averaging set to $K=1$. Here $\R'_{l}$ is the sample covariance matrix for the $l$th subarray, and $\breve{\R}$ is the average of $N_K$ of these. Note that instead of performing the temporal sum last as here, one could take more temporal samples into consideration in the computation of each $r_{ij}$.}\label{mvdr_build_R}
\end{figure}

@@

\subsection[Computing the Spatial Covariance Matrix]{Computing the Spatial Covariance Matrix, $\hat{\boldsymbol{R}}$}\label{computing_eR}

% As Fig. \ref{mvdr_complexity} illustrates the computation of $\eR$

As observed in Fig. \ref{mvdr_complexity}, a direct computation of $\eR$ by implementing (\ref{spatialR}) and (\ref{finalR}) is the greatest computational burden. We target this issue by first minimizing arithmetic operations, then aim to get as close to the peak arithmetic throughput on the GPU as possible.

@@

In Fig. \ref{mvdr_build_R}, we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$, and temporal averaging is set to $K=1$. Here the spatial sum is carried out before the temporal sum. To reduce arithmetic operations we may:
%
\begin{enumerate}
\item exploit the fact that $\eR$ is Hermitian positive semidefinite to compute only one half of it;
\item avoid redundant multiplications;
\item avoid redundant summations by first computing the upper row of $\eR$ and then adding/subtracting to these iteratively to find the diagonals of $\eR$.
\end{enumerate}

\ifPhdDoc
\begin{figure}[tp]\centering
\includegraphics[width=\linewidth]{gfx/buske5\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske5\figPostfix.eps} 
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/buildR-breakdown.eps}
\fi\fi%
\caption{Arithmetic optimization of computing $\eR$: Relative reduction in arithmetic complexity compared to the initial implementation shown in Fig. \ref{mvdr_complexity} (higher is better) . Note how the arithmetic count is reduced by a factor 4-10 in the memory-optimized case, and by a factor 6-22 in the instruction-optimized case.}\label{mvdr_complexity_optimized}
\end{figure}

To study the effect of these optimizations we altered the complexity formulas to take them into account. Exploiting symmetry and performing iterative summing (step 1 and 3) is always desirable, but avoiding redundant multiplications (step 2) comes at the cost of increased memory consumption. When all optimizations are applied, we call it a "minimum instructions" solution, and when step 2 is left out, we call it a "minimum memory" solution. Fig. \ref{mvdr_complexity_optimized} compares the two solutions, and formulas may be found in Appendix \ref{mvdr_formulas}. Here the complexity curves are relative to the reference implementation in Fig. \ref{mvdr_complexity}. Both solutions reduce the complexity considerably, and recomputing the multiplications does not make that much of a difference. We selected the memory minimized version since our algorithm is severely memory bound, as we will see next.

\ifPhdDoc
\begin{figure}[p]\centering
\includegraphics[width=.997\linewidth]{gfx/buske6\figPostfix.eps}
% Caption is the same, but using newlines for the items in it (since it is now on a page)
\caption{MVDR implemented on a GPU. We do this in 3 steps, where each step process the full image before moving on to the next step. \\[.5\baselineskip]\textbf{Step~1}: The sample covariance matrix $\eR$ is formed by threads running along its diagonals. This allows spatial averaging to be implemented in a computationally efficient manner and minimizes interthread communication.
\\[.5\baselineskip]\textbf{Step~2}: $\eRi\1$ is computed using the heavily optimized batched linear equation solver from Nvidia~\cite{Nvidia}. 
\\[.5\baselineskip]\textbf{Step~3}: The beamformer output $z$ is computed in a straightforward fashion using $L$ threads that first sum the subarrays and then apply the MVDR weighting function. A single thread finally sums all the channels up.}\label{mvdr_implementation}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske6\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_implementation.eps}
\fi
\caption{MVDR implemented on a GPU. We do this in 3 steps, where each step process the full image before moving on to the next step. \emph{Step~1}: The sample covariance matrix $\eR$ is formed by threads running along its diagonals. This allows spatial averaging to be implemented in a computationally efficient manner and minimizes interthread communication. \emph{Step~2}: $\eRi\1$ is computed using the heavily optimized batched linear equation solver from Nvidia~\cite{Nvidia}. \emph{Step~3}: The beamformer output $z$ is computed in a straightforward fashion using $L$ threads that first sum the subarrays and then apply the MVDR weighting function. A single thread finally sums all the channels up.}\label{mvdr_implementation}
\fi%
\end{figure}

@@

When we compute $\eR$, we very rarely perform more than 1-3 floating point operations for every float read or written to memory. Unfortunately, the GPU prefers kernels to be more arithmetically intensive than this. This can be inferred from Table \ref{throughputs}, %
%
\begin{table}[!b]\centering%\normalsize
\begin{tabular}[c]{l r r r@{}  l}\hline
\rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{arith}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$/B$_\text{arith}$} \\\hline
Arithmetic & 1.03 Tflop/s & & &\\
Global memory & & 36 Gfloats/s & \hspace{30pt} 1 &:30 \\
Shared memory & & 257 Gfloats/s & 1 &:4 \\
Registers & & $>$1.5 Tfloats/s & $>$3 &:2~\cite{Vasilyy}
\end{tabular}
\caption{Nvidia Quadro 6000: Memory throughput, $B_{\lowercase{\text{mem}}}$, compared to arithmetic throughput, $B_\text{arith}$.}\label{throughputs}
\end{table}%
%
where the peak bandwidth of the three types of GPU memory are compared to the peak arithmetic throughput (see appendix \ref{throughput} for derivations). Global memory (RAM), in which all data must reside at some point, is at best only able to move one float for every 30 floats processed by the CUDA cores, and potentially much worse if care is not taken to use "coalesced" memory access patterns (that avoid memory bank conflicts). This is why the usage of global memory should be minimized. Shared memory, on the other hand, is a very fast level 1 cache that is shared by all threads on an SM. At peak performance, it is almost able to keep up with the computing units, but to obtain this performance we must make sure that:
\begin{enumerate}
\item the data needed to compute a pixel can fit in shared memory, while
\item the access patterns we use promote maximum bandwidth.
\end{enumerate}
 %Further alleviation is also possible by exposing instruction level parallelism within a thread, as this promotes the use of registers with a bandwidth at least 6 times higher than that of shared memory~\cite{Vasilyy}\todo{so why don't we? rephrase}.

@@

In Fig. \ref{mvdr_implementation}, we illustrate how we get around these challenges. First we make each SM compute entire range lines of pixels, then load the subset of $\x$ that all these pixels depend upon into shared memory. This lets us perform temporal averaging without having to read additional channel data.

@@

The beamformer output $z$ is computed on a per-pixel basis by first computing the MVDR weights $\w$, which are merely scaled versions of $\eRi\1$ (\ref{weights}). Then the sum in (\ref{finalZ}) is found by assigning a group of $L$ threads to respective elements of $x_l$, which then proceed to accumulate these for all $N_L$ subarrays. The resulting data vector is finally multiplied with the weight vector using the same threads, and then a single thread is used to sum these products to obtain $z$.

>>>>

This article is outlined as follows: In Section \ref{I_methods}, we introduce the concept of adaptive beamforming and provide details on the MVDR method. Then, in \ref{I_maptogpu}, we investigate the complexity issue, discuss means for reducing arithmetic complexity, and detail an implementation that makes efficient use of the GPU's parallel resources. The final design is assessed in Sections \ref{I_images_and_benchmarks} and \ref{I_discussion}, where we provide benchmarks, comparisons with similar CPU implementations, and measures of how efficiently our implementation makes use of the GPU's resources.

@@

\section{Methods}\label{I_methods}

@@

\subsection{Beamforming}
\ifPhdDoc
\begin{figure}[htbp]\centering
\includegraphics[width=\linewidth]{gfx/buske1\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[htbp]\centering
\includegraphics[width=0.8\linewidth]{gfx/buske1\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/beamforming.eps}
\fi\fi%
\caption{Beamforming principle. Signal signature is first removed by matched filtering. Then, before summation, a suitable set of delays $\Delta$ and weights $w$ are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{I_beamforming}
\end{figure}
Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter (Fig. \ref{I_beamforming}). Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$. To simplify notation, we make the dependence on $\theta$ implicit from now on.

By definition, the beamformer output $z[n]$ can now be expressed as the weighted sum of all the delayed data samples:
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{I_z}
\end{align}
where $w_m$ is the weight factor assigned to channel $m$. With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer. A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.

Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield. In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality. The MVDR beamformer is one such method. It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}. This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{I_weights}
\end{gather}
where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. Since we pre-steer our data to every pixel in the image we simplify (\ref{I_weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering. To estimate $\R$ we compute a sample covariance matrix $\eR$. In this computation we perform some degree of:
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
\end{itemize}%
%
% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009a}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}. 
%
Combined, these steps will also ensure a numerically well conditioned $\eR$.

We perform \emph{temporal} and \emph{spatial averaging} first and put the result in an intermediate sample covariance matrix $\breve{\R}$. To do this we need to segment our array into subarrays. If we let $x_l[n]$ represent the data vector from subarray $l$ by
\begin{gather}
\x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T,
\end{gather}
then $\breve{\R}$ can be calculated as
\begin{gather}
\breve{\R}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n']\x_l\H[n'] \in\mathbb{C}^{L,L},\label{I_spatialR}
\end{gather}
where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.

The final estimate $\eR$ is found by adding a fraction $d$ of the total power of $\breve{\R}[n]$ to its diagonal~\cite{Synnevag2007}:
\begin{align}
\eR[n] = \breve{\R}[n] + \I \frac{d}{L} \tr\{\breve{\R}[n]\},\label{I_finalR}
\end{align}
where $\I$ is an identity matrix, $\tr\{\cdot\}$ represents the matrix trace operation, and $\tr\{\breve{\R}[n]\}$ is an estimate of the energy received from this pixel.

Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{M,M}$ to $\mathbb{C}^{L,L}$, and hence will produce an $L$-element weight set when substituted into (\ref{I_z}). This weight set is applied to all the subarrays, before computing the beamformer output as in (\ref{I_z}). Or, equivalently, we may apply the weight set to the sum of all the subarrays by
%
\begin{align}
z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{I_finalZ}
\end{align}
%
\ifPhdDoc
\begin{figure}[tbp]\centering
\includegraphics[width=\linewidth]{gfx/buske2\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske2\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/implementation.eps}
\fi\fi%
\caption{MVDR beamforming. For each pixel in range and azimuth:\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed; \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver;\newline
3. and the beamformer output $z$ is computed from (\ref{I_finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{I_weights}). } \label{I_mvdr_beamforming}
\end{figure}
As summarized in Fig. \ref{I_mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
\begin{enumerate}
\item computing the sample covariance matrix $\eR$ in (\ref{I_finalR});
\item computing $\eRi\1$ in (\ref{I_weights});
\item computing the beamformer output $z$ in (\ref{I_finalZ}).
\end{enumerate}
Next we will evaluate these steps in terms of arithmetic complexity, and then discuss their mappability to parallel hardware.

@@

If we neglect spatial and temporal averaging, then the computation of $\eR$ is reduced to a single outer product with complexity of O($M^2$), and the inversion is then of O($M^3$). This might lead us to believe that the inversion step is by far the most complex. But if we implement spatial and temporal averaging as in (\ref{I_spatialR}), then computing $\eR$ is of O($N_K N_L L^2$) and the inversion of O($L^3$). Computing $\eR$ is now the most complex operation whenever $N_LN_K>L$.
%
\ifPhdDoc
\begin{figure}[tbp]\centering
\includegraphics[width=\linewidth]{gfx/buske3\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske3\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_complexity.eps}
\fi\fi%
\caption{Per-pixel computational complexity of the steps in MVDR beamforming (before any optimizations). To avoid signal cancellation in an active sonar system we usually set $L<\frac{M}{2}$, in which region the computation of $\eR$ dominates in terms of arithmetic complexity, especially when performing temporal averaging.}\label{I_mvdr_complexity}
\end{figure}
To visualize these relations, and include the effects of using complex numbers, we set up complexity formulas that account for the total number of arithmetic operations for each step in the MVDR process (see appendix \ref{I_mvdr_formulas}). The only excluded operation was partial pivoting used in the inversion step, but this should contribute marginally to the end result. The entire range of possible subarray sizes from $L\in[1,M]$ was finally evaluated, with temporal averaging set to $K\in\{0,1,2\}$, and the number of channels set to $M\in\{8,16,32\}$. The results are shown in \Fig{I_mvdr_complexity}.

@@

\section{Mapping the MVDR to a GPU}\label{I_maptogpu}

@@

To use Nvidia's own terminology, the Quadro 6000 comprises 14 streaming multiprocessors (SMs), each having 32 CUDA cores that execute a common program called a kernel. Combined these cores deliver a peak performance of more than 1\;Tflop/s (appendix \ref{I_throughput}). In practice this performance is hard to obtain, but one can get fairly close by balancing the load evenly on all cores, and by trying to avoid that some cores are forced to idle due to a pending data transfer or thread synchronization. As the steps in the MVDR method require different strategies for achieving this, we have designed a different kernel and configuration for each of them. We will pay particular attention to building $\eR$, since the potential for gaining overall speedups are greatest here.

@@

\ifPhdDoc
\begin{figure}[tbp]\centering
\includegraphics[width=\linewidth]{gfx/buske4\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske4\figPostfix.eps}
\else
\begin{figure}[t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_build_R.eps}
\fi\fi%
\caption{Step 1: Building $\eR$. This is a visualization of how $\eR$ could be built in a case with $M=5$ sensors, with subarray size $L=3$ and temporal averaging set to $K=1$. Here $\R'_{l}$ is the sample covariance matrix for the $l$th subarray, and $\breve{\R}$ is the average of $N_K$ of these. Note that instead of performing the temporal sum last as here, one could take more temporal samples into consideration in the computation of each $r_{ij}$.}\label{I_mvdr_build_R}
\end{figure}

@@

\subsection[Computing the Spatial Covariance Matrix]{Computing the Spatial Covariance Matrix, $\hat{\boldsymbol{R}}$}\label{I_computing_eR}

% As Fig. \ref{I_mvdr_complexity} illustrates the computation of $\eR$

As observed in Fig. \ref{I_mvdr_complexity}, a direct computation of $\eR$ by implementing (\ref{I_spatialR}) and (\ref{I_finalR}) is the greatest computational burden. We target this issue by first minimizing arithmetic operations, then aim to get as close to the peak arithmetic throughput on the GPU as possible.

@@

In Fig. \ref{I_mvdr_build_R}, we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$, and temporal averaging is set to $K=1$. Here the spatial sum is carried out before the temporal sum. To reduce arithmetic operations we may:
%
\begin{enumerate}
\item exploit the fact that $\eR$ is Hermitian positive semidefinite to compute only one half of it;
\item avoid redundant multiplications;
\item avoid redundant summations by first computing the upper row of $\eR$ and then adding/subtracting to these iteratively to find the diagonals of $\eR$.
\end{enumerate}

\ifPhdDoc
\begin{figure}[tp]\centering
\includegraphics[width=\linewidth]{gfx/buske5\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske5\figPostfix.eps} 
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/buildR-breakdown.eps}
\fi\fi%
\caption{Arithmetic optimization of computing $\eR$: Relative reduction in arithmetic complexity compared to the initial implementation shown in Fig. \ref{I_mvdr_complexity} (higher is better) . Note how the arithmetic count is reduced by a factor 4-10 in the memory-optimized case, and by a factor 6-22 in the instruction-optimized case.}\label{I_mvdr_complexity_optimized}
\end{figure}

To study the effect of these optimizations we altered the complexity formulas to take them into account. Exploiting symmetry and performing iterative summing (step 1 and 3) is always desirable, but avoiding redundant multiplications (step 2) comes at the cost of increased memory consumption. When all optimizations are applied, we call it a "minimum instructions" solution, and when step 2 is left out, we call it a "minimum memory" solution. Fig. \ref{I_mvdr_complexity_optimized} compares the two solutions, and formulas may be found in Appendix \ref{I_mvdr_formulas}. Here the complexity curves are relative to the reference implementation in Fig. \ref{I_mvdr_complexity}. Both solutions reduce the complexity considerably, and recomputing the multiplications does not make that much of a difference. We selected the memory minimized version since our algorithm is severely memory bound, as we will see next.

\ifPhdDoc
\begin{figure}[p]\centering
\includegraphics[width=.997\linewidth]{gfx/buske6\figPostfix.eps}
% Caption is the same, but using newlines for the items in it (since it is now on a page)
\caption{MVDR implemented on a GPU. We do this in 3 steps, where each step process the full image before moving on to the next step. \\[.5\baselineskip]\textbf{Step~1}: The sample covariance matrix $\eR$ is formed by threads running along its diagonals. This allows spatial averaging to be implemented in a computationally efficient manner and minimizes interthread communication.
\\[.5\baselineskip]\textbf{Step~2}: $\eRi\1$ is computed using the heavily optimized batched linear equation solver from Nvidia~\cite{Nvidia}. 
\\[.5\baselineskip]\textbf{Step~3}: The beamformer output $z$ is computed in a straightforward fashion using $L$ threads that first sum the subarrays and then apply the MVDR weighting function. A single thread finally sums all the channels up.}\label{I_mvdr_implementation}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske6\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_implementation.eps}
\fi
\caption{MVDR implemented on a GPU. We do this in 3 steps, where each step process the full image before moving on to the next step. \emph{Step~1}: The sample covariance matrix $\eR$ is formed by threads running along its diagonals. This allows spatial averaging to be implemented in a computationally efficient manner and minimizes interthread communication. \emph{Step~2}: $\eRi\1$ is computed using the heavily optimized batched linear equation solver from Nvidia~\cite{Nvidia}. \emph{Step~3}: The beamformer output $z$ is computed in a straightforward fashion using $L$ threads that first sum the subarrays and then apply the MVDR weighting function. A single thread finally sums all the channels up.}\label{I_mvdr_implementation}
\fi%
\end{figure}

@@

When we compute $\eR$, we very rarely perform more than 1-3 floating point operations for every float read or written to memory. Unfortunately, the GPU prefers kernels to be more arithmetically intensive than this. This can be inferred from Table \ref{I_throughputs}, %
%
\begin{table}[!b]\centering%\normalsize
\begin{tabular}[c]{l r r r@{}  l}\hline
\rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{arith}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$/B$_\text{arith}$} \\\hline
Arithmetic & 1.03 Tflop/s & & &\\
Global memory & & 36 Gfloats/s & \hspace{30pt} 1 &:30 \\
Shared memory & & 257 Gfloats/s & 1 &:4 \\
Registers & & $>$1.5 Tfloats/s & $>$3 &:2~\cite{Vasilyy}
\end{tabular}
\caption{Nvidia Quadro 6000: Memory throughput, $B_{\lowercase{\text{mem}}}$, compared to arithmetic throughput, $B_\text{arith}$.}\label{I_throughputs}
\end{table}%
%
where the peak bandwidth of the three types of GPU memory are compared to the peak arithmetic throughput (see appendix \ref{I_throughput} for derivations). Global memory (RAM), in which all data must reside at some point, is at best only able to move one float for every 30 floats processed by the CUDA cores, and potentially much worse if care is not taken to use "coalesced" memory access patterns (that avoid memory bank conflicts). This is why the usage of global memory should be minimized. Shared memory, on the other hand, is a very fast level 1 cache that is shared by all threads on an SM. At peak performance, it is almost able to keep up with the computing units, but to obtain this performance we must make sure that:
\begin{enumerate}
\item the data needed to compute a pixel can fit in shared memory, while
\item the access patterns we use promote maximum bandwidth.
\end{enumerate}
 %Further alleviation is also possible by exposing instruction level parallelism within a thread, as this promotes the use of registers with a bandwidth at least 6 times higher than that of shared memory~\cite{Vasilyy}\todo{so why don't we? rephrase}.

@@

In Fig. \ref{I_mvdr_implementation}, we illustrate how we get around these challenges. First we make each SM compute entire range lines of pixels, then load the subset of $\x$ that all these pixels depend upon into shared memory. This lets us perform temporal averaging without having to read additional channel data.

@@

The beamformer output $z$ is computed on a per-pixel basis by first computing the MVDR weights $\w$, which are merely scaled versions of $\eRi\1$ (\ref{I_weights}). Then the sum in (\ref{I_finalZ}) is found by assigning a group of $L$ threads to respective elements of $x_l$, which then proceed to accumulate these for all $N_L$ subarrays. The resulting data vector is finally multiplied with the weight vector using the same threads, and then a single thread is used to sum these products to obtain $z$.
